---
title: "Flow Plots: Where the Movement Goes"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Flow Plots: Where the Movement Goes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
has_cograph <- requireNamespace("cograph", quietly = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7.5,
  fig.height = 4.6,
  dpi = 120,
  eval = has_cograph
)
```

A state distribution that stays flat across time is ambiguous evidence. It
is equally consistent with a cohort in which nobody moves and a cohort in
which everybody swaps places, because both leave the same number of
students in each state at each moment. The engagement data turn out to be
the second kind: the share of disengaged students stays between a fifth
and a quarter of the cohort at every one of the eight course positions,
yet almost a third of consecutive observations are a change of state.
Nothing in a
distribution plot, and nothing in a sequence plot sorted by similarity,
separates those two worlds.

`flow_plot()` closes that gap. It draws the movement itself — how many
students leave each state, and which state they arrive in — using
[cograph](https://github.com/mohsaqr/cograph) for the rendering, and
`transition_plot()` collapses the same movement into a network whose node
sizes come from `Nestimate`'s transition-network centralities. This
vignette is the third in the series: `vignette("get-started")` covers the
interface, `vignette("vasstra-tutorial")` runs the complete analysis, and
this one covers the views that answer *where movement goes*, in this
order:

1. Why flows need their own view.
2. Aggregated flows: the alluvial.
3. Individual flows: one line per student.
4. Flows inside a single trajectory.
5. Controlling colour, bundling, and layout.
6. The transition network: movement collapsed over time.

| Question | View |
|---|---|
| What is the mix at each time? | `plot(fit, which = "sequences", type = "distribution")` |
| Who resembles whom? | `plot(fit, which = "sequences")` (heatmap) |
| Where does movement go? | `flow_plot(fit)` |
| Which students move where? | `flow_plot(fit, type = "individual")` |
| How does one trajectory move? | `flow_plot(fit, group = ...)` |
| Which states attract movement overall? | `transition_plot(fit)` |
| The same, as numbers | `transition_centrality(fit)` |

```{r fit}
library(VaSStra)
data("engagement", package = "VaSStra")

fit <- vasstra(
  engagement,
  state_labels = c("Disengaged", "Average", "Active"),
  dissimilarity = "lcs",
  cluster_method = "ward.D2",
  trajectory_labels = c(
    "Mostly active",
    "Mostly average",
    "Mostly disengaged"
  ),
  positive_states = "Active",
  negative_states = "Disengaged"
)
```

## 1. Why flows need their own view

The transition table is the numeric form of the same information, and it
already shows why the flat distribution is misleading.

```{r transitions}
summary(fit$sequences)
fit$sequences$transitions
```

Two facts drive everything below. First, persistence is high but far from
total: 696 of the 994 consecutive pairs stay put, so 30 percent are moves.
A distribution plot renders those 298 moves as a flat band, because they
cancel — 73 students leave Disengaged for Average while 73 travel the
other way, and 72 move up from Average to Active against 73 coming back
down. The composition is stable precisely *because* the exchange is
balanced, which is a different claim from stability and has different
consequences for anyone designing an intervention.

Second, the corners of the table are nearly empty. Only 5 of 994
transitions run from Disengaged straight to Active, and only 2 run the
other way. Movement between the extremes essentially always passes
through Average, which makes the middle state a waypoint rather than
merely an intermediate label. That is a structural claim about the state
system, and it is the kind of thing the eye reads instantly from a flow
diagram and slowly from a nine-row table.

## 2. Aggregated flows: the alluvial

`flow_plot()` draws one band per transition, with band width proportional
to the number of students making it, at every consecutive pair of time
points at once.

```{r alluvial}
flow_plot(fit)
```

The near-absent corner transitions appear as the missing bands: only
hair-thin threads cross the full height of the figure between the top
block and the bottom one, against the substantial ribbons joining each
block to its neighbour. The balanced exchange appears as the criss-cross
between neighbouring
states, which is thick in both directions at every step and leaves the
block heights almost unchanged from column to column. Read together with
the transition table, the picture supplies the shape and the table
supplies the counts.

Bands are coloured by the state they leave, which is what makes the
outflow of a single state followable across the whole figure. Colouring by
destination answers the mirrored question — which state a given block
recruits from — and is one argument away.

```{r alluvial-destination}
flow_plot(fit, color_by = "destination")
```

The colours, the state order, and the time labels come from the fit, so
this figure and the sequence heatmap from the tutorial can be placed side
by side without re-reading a legend.

## 3. Individual flows: one line per student

Aggregated bands merge students who move together. The individual view
keeps them apart, drawing one line per student coloured by the state they
started in, which turns the question from *how much flow* into *whose*.

```{r individual}
flow_plot(fit, type = "individual")
```

Colouring by first state makes early position followable to the end: the
green lines beginning in Active stay largely within the lower band, while
the orange lines that begin in Average are the ones ranging across the
full height of the figure. This is the visual counterpart of the trajectory
description's mobility columns — the Average-heavy group records 2.33
transitions per student against 1.83 and 1.89 for the two extreme groups,
and here that difference is the visible fanning of the orange lines.

With 142 students the lines would overplot, so `flow_plot()` bundles them
by default and annotates the figure with how many students one line
represents. Setting `bundle = FALSE` draws every student, which is worth
doing for small cohorts and rarely worth doing for large ones.

```{r individual-last}
flow_plot(fit, type = "individual", color_by = "last")
```

Colouring by the *final* state instead reverses the reading: it traces
where each of today's groups came from, which is the useful direction when
the outcome is known and the question is what preceded it.

## 4. Flows inside a single trajectory

The trajectories are defined by whole-sequence similarity, so the natural
follow-up is whether a trajectory is internally uniform or merely averages
to its label. Flow diagrams draw a single panel and cannot be faceted, so
`group` restricts the plot to one trajectory's students.

```{r group-average}
flow_plot(fit, group = "Mostly average", type = "individual")
```

The label survives the check: the group is overwhelmingly a single orange
band, with a thin blue thread of excursions into Disengaged and a small
Active block that never grows. Its mean within-trajectory distance is the
largest of the three (4.96 against 4.26 and 3.86), and this is what that
looseness looks like — not a mixture of different journeys, but one
journey with the most frequent small departures from it.

```{r group-disengaged}
flow_plot(fit, group = "Mostly disengaged")
```

The most disengaged group is the tightest, and its aggregated flows show
why: the Disengaged block absorbs nearly everything at every step, with
its negative exposure of 0.82 and integrative potential of exactly zero
meaning no student in this group is in the Active state at the end. The
contrast with the previous figure is the substantive result — these
trajectories differ in their movement, not only in their average level.

```{r trajectory-table}
as.data.frame(fit, unit = "trajectory")
```

## 5. Controlling colour, bundling, and layout

`flow_plot()` sets the palette, the state order, the time labels, and a
readable geometry, and passes anything else through to cograph, so a
figure destined for a manuscript can be tuned without abandoning the
wrapper.

```{r tuned}
flow_plot(
  fit,
  main = "Engagement flows across eight course positions",
  colors = c("#B3B3B3", "#4C9BE8", "#0B5394"),
  flow_alpha = 0.75,
  threshold = 3
)
```

`colors` takes one colour per state in state order and applies to
everything the figure draws, so a custom palette stays consistent between
nodes and bands — here an ordered grey-to-navy ramp, which suits an ordered
state system better than the categorical default.

`threshold` suppresses flows below a given number of students, and this
data set makes the choice unusually easy to defend. Counted step by step,
the corner transitions between Disengaged and Active never exceed two
students, while the smallest transition between neighbouring states is
five. A threshold of three therefore removes every corner transition and
retains every neighbouring one, which is why the criss-cross survives
intact above while the hair-thin diagonals are gone. Thresholds chosen
without that check will quietly delete real movement, so it is worth
reading the transition table before setting one.

Two arguments are specific to the individual view. `bundle` sets how many
students one line represents — `"auto"` by default, `FALSE` for every
student, or a number — and `bundle_max` sets the number of lines that
triggers automatic bundling.

```{r bundle}
flow_plot(
  fit,
  group = "Mostly disengaged",
  type = "individual",
  bundle = FALSE
)
```

## 6. The transition network: movement collapsed over time

Every view so far keeps time on the horizontal axis. Collapsing all seven
steps into one network drops the timing and asks a different question:
across the whole course, which states attract movement and which emit it?
`transition_plot()` builds that network with `Nestimate::build_tna()`,
takes its centralities from `Nestimate::net_centrality()`, and hands the
result to `cograph::splot()`, which recognises the network and supplies
the TNA styling, the labels, and the rings showing each state's initial
probability. Node size is in-strength — the total incoming transition
weight — so the biggest node is the state the cohort most often moves
into.

```{r transition, fig.height = 5.5}
transition_plot(fit)
```

The same centralities are available as a table without drawing anything,
which is what makes the size comparison checkable rather than
impressionistic.

```{r transition-centrality}
transition_centrality(fit)
```

Average's in-strength of 0.57 is more than three times either extreme, and
that single number is the quantitative form of the waypoint result from
section 1: movement does not merely pass through Average, it accumulates
there. The two extremes are almost indistinguishable on this measure
(0.159 against 0.171) even though their self-transition probabilities
differ noticeably (0.67 for Disengaged against 0.73 for Active), because
in-strength excludes self-loops by default. Persistence and attraction are
genuinely different properties, and keeping them apart is the reason for
that default.

Setting `loops = TRUE` folds persistence back in, and the ordering changes
character entirely — the nodes become nearly equal in size because every
state mostly retains its own members.

```{r transition-loops, fig.height = 5.5}
transition_plot(fit, loops = TRUE)
```

`weights = "count"` swaps transition probabilities for raw counts, which
is the version to report when the absolute volume of movement matters
rather than its conditional likelihood, and `size` accepts any measure
`Nestimate::net_centrality()` offers — `measures = "all"` on
`transition_centrality()` lists them.

```{r transition-counts, fig.height = 5.5}
transition_plot(fit, weights = "count", size = "OutStrength")
```

Restricted to one trajectory, the network becomes a compact summary of how
that group moves — and a group that never reaches a state simply has no
node for it.

```{r transition-group, fig.height = 5.5}
transition_plot(fit, group = "Mostly disengaged")
```

### Sequences and network side by side

The network summarises movement but hides the data it came from, so the
conventional presentation puts the two together: the raw sequences on the
left, their transition network on the right. `sequences = TRUE` draws both
on one device, and because the panels share a palette a state has the same
colour in each.

```{r transition-paired, fig.width = 10, fig.height = 4.4}
transition_plot(fit, sequences = TRUE, group = "Mostly active")
```

The pairing is what makes each panel legible. The left panel shows a green
field with orange interruptions, and the right explains its shape: Active
retains 0.84 of its members from one position to the next, and the only
substantial route out of it leads to Average rather than to Disengaged,
whose node is barely present. Reading either panel alone would leave the
other half of that statement unsupported.

`sequences` also accepts `"heatmap"` or `"distribution"` for the left
panel, and the argument changes only the layout — the centralities are
identical either way.

```{r transition-paired-contrast, fig.width = 10, fig.height = 4.4}
transition_plot(fit, sequences = TRUE, group = "Mostly disengaged")
```

## Where flow plots fit

The package keeps two rendering paths, and each owns a distinct question.
Nestimate renders the sequence views — heatmap, index, and distribution —
which are all organised by *subject similarity or composition*, and it also
builds the transition network and its centralities. cograph draws the
movement views: the flow diagrams, organised by *movement between
consecutive times*, and the transition network, which discards timing
altogether.

The three answers compose. The distribution establishes that the
composition is stable; the alluvial establishes that the stability is an
equilibrium rather than an absence of change; and the transition network
puts a number on where that exchange concentrates. None of the three is a
substitute for the others, and the first on its own is the one most easily
misread.

Method overview:
[VaSSTra chapter](https://lamethods.org/book1/chapters/ch11-vasstra/ch11-vasstra.html).
