---
title: "VaSSTra in Four Simple Steps"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{VaSSTra in Four Simple Steps}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

VaSSTra turns repeated multivariate measurements into states, state
sequences, and groups of similar trajectories. The whole analysis is one
call, and every part of it can also be run as an explicit step.

```{r data}
library(VaSSTra)

set.seed(2026)
data <- expand.grid(
  student = seq_len(36),
  course = seq_len(6)
)
trajectory <- ceiling(data$student / 12)
# Three groups of twelve students climb the same three-phase ladder,
# each starting one phase further along, so the groups differ in timing
# rather than in level.
phase <- ceiling(data$course / 2)
latent_state <- (phase + trajectory - 2) %% 3 + 1
data$views <- latent_state * 6 + rnorm(nrow(data), sd = 0.08)
data$sessions <- latent_state * 3 + rnorm(nrow(data), sd = 0.08)
data$duration <- latent_state * 12 + rnorm(nrow(data), sd = 0.08)
```

## One call

`vasstra()` detects the subject, time, and indicator roles from common
column names (or attached role metadata), compares two through six states
and trajectories, and fits the recommended counts. Every automated decision
is reported as it is made and recorded inside the fit.

```{r one-call}
fit <- vasstra(data)
fit
```

Tidy tables come straight from the fit at the analysis unit you need:

```{r tidy-units}
head(as.data.frame(fit))
head(as.data.frame(fit, unit = "observation"))
as.data.frame(fit, unit = "trajectory")
```

Name any decision to take it over; automation only fills what you omit.
Labels are the shortest way to fix a count — three labels mean three
states — and a candidate vector such as `n_states = 2:4` compares exactly
those counts and fits the recommended one.

```{r explicit-call, eval = FALSE}
fit <- vasstra(
  data,
  state_labels = c("Low", "Average", "High"),
  trajectory_labels = c("Stable low", "Moving", "Stable high"),
  dissimilarity = "lcs",
  cluster_method = "ward.D2",
  positive_states = "High",
  negative_states = "Low"
)
```

## Evaluate the clustering

`evaluate()` compares the fitted numbers of states and trajectories against
the neighboring counts and reports per-cluster quality. Printing shows both
tables; plotting draws the selection curve, per-cluster silhouette widths,
and group sizes.

```{r evaluate}
evaluation <- evaluate(fit)
evaluation
```

```{r evaluate-plot, fig.height = 6}
plot(evaluation)
```

The same verb works on a single step, with any candidate range:

```{r evaluate-states}
state_evaluation <- evaluate(fit$states, n_states = 2:5)
as.data.frame(state_evaluation)
summary(state_evaluation)
```

## Tidy fit indices

`fit_indices()` reports the fit statistics of the selected clustering as
one tidy row, or one row per compared candidate with `compare = TRUE`.
LPA models report the complete information-criterion family together
with entropy and the extreme average posterior class probabilities;
hard-clustering methods report their own objectives instead, and columns
that do not apply are dropped.

```{r fit-indices}
fit_indices(fit)
fit_indices(fit, compare = TRUE)
fit_indices(fit, step = "trajectories", compare = TRUE)
```

## Name groups after inspecting them

`set_labels()` renames fitted states or trajectories in place — nothing
is refitted, no value changes — and the new names flow through every
table, sequence, and plot. Rename everything with a full vector, or only
some groups by name.

```{r set-labels}
fit <- set_labels(
  fit,
  states = c("Low", "Average", "High"),
  trajectories = c("Mostly high", "Mixed", "Mostly low")
)
summary(fit)
```

## The same analysis in four explicit steps

Each step accepts the result of the previous step and also runs alone with
the same automated defaults. Use the steps when intermediate results are
needed, or when states come from another method.

```{r steps}
states <- step1_states(data, n_states = 3,
                       labels = c("Low", "Average", "High"))
sequences <- step2_sequences(states)
trajectories <- step3_trajectories(sequences, n_trajectories = 3)
description <- step4_describe(
  trajectories,
  positive_states = "High",
  negative_states = "Low"
)
summary(description)
```

Existing states from LPA, LCA, or any other source enter at step 2: pass a
data frame whose state column is detected (or named with `state`).

## State plots

State results support profile, bar, heatmap, and size plots without adding
a plotting dependency, plus a combined overview.

```{r state-plots, fig.height = 6}
plot(states, type = "all")
```

```{r state-profile}
plot(states, type = "profile")
```

## Sequence and trajectory plots

Every plot containing state sequences is rendered by Nestimate:

```{r sequence-plot}
plot(trajectories, type = "index")
plot(trajectories, type = "distribution")
```

## Compare candidates in full detail

When automation should be replaced by an explicit comparison,
`state_choices()` and `trajectory_choices()` fit a grid across methods and
counts and return one tidy row per candidate. Nothing is selected silently;
pass the inspected `candidate_id` to fit.

```{r state-choices}
state_options <- state_choices(
  data,
  n_states = 2:4,
  method = c("kmeans", "pam", "ward.D2")
)
state_options
plot(state_options, metric = "silhouette")
chosen_states <- fit_state_choice(
  state_options,
  n_states = 3,
  method = "kmeans",
  labels = c("Low", "Average", "High")
)
```

`fit_state_choice(state_options)` with no selection fits the recommended
candidate; `candidate_id` remains available for exact row selection.

Average silhouette is the common diagnostic across all methods; larger
values indicate better-separated groups. Recommendations identify the best
eligible candidate separately for each method rather than declaring one
algorithm universally best. LPA candidates are recommended by conventional
BIC, with AIC, silhouette, or native mclust ICL available through
`lpa_criterion`.

```{r trajectory-choices}
trajectory_options <- trajectory_choices(
  step2_sequences(chosen_states),
  n_trajectories = 2:4,
  dissimilarity = c("hamming", "lcs"),
  method = c("pam", "ward.D2")
)
trajectory_options
plot(trajectory_options, metric = "silhouette")
```

Recommendations are made within each distance-method pair because mean
within-trajectory distances are only comparable within one dissimilarity.

States default to mclust latent profiles (`"EEI"`, tidyLPA model 1);
base R supplies k-means and hierarchical clustering, `cluster` supplies
PAM, and Nestimate supplies sequence distances, trajectory clustering,
and plots. No TraMineR plotting path or dependency is required.
