---
title: "Plotting the results"
output:
  html_document:
    toc: true
---

```{r setup, include = FALSE}
# cograph is a suggested package; its flow and transition chunks are skipped
# when it is not installed, so the vignette still builds.
has_cograph <- requireNamespace("cograph", quietly = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = TRUE,
  fig.width = 7.5,
  fig.height = 6,
  dpi = 120,
  message = TRUE,
  warning = TRUE
)
```

This review exercises every VaSSTra plot on the longitudinal engagement
data from
[Chapter 11 of *Learning Analytics Methods and Tutorials*](https://lamethods.org/book1/chapters/ch11-vasstra/ch11-vasstra.html):
the state, sequence, and trajectory views (base graphics, no `ggplot2`
dependency), the combined per-trajectory grid, the `evaluate()` panels,
and the `cograph`-rendered flow and transition networks. Every view shares
one state palette and one state order, both of which can be pinned on the
fit with `state_colors` and `state_order`.

## The whole analysis in one call

`vasstra(engagement)` needs nothing else. Roles come from the attached
metadata, and the state and trajectory counts are selected by comparing
two through six candidates. Here BIC favours five states, so the fully
automatic fit reports five; the chapter analysis below fixes three for
interpretability. Each decision is reported as a message and recorded in
the fit.

```{r one-call}
library(VaSSTra)
data("engagement", package = "VaSSTra")

fit_auto <- vasstra(engagement)
fit_auto
```

## Evaluate the automatic clustering

`evaluate()` returns the candidate comparison with `best` and `fitted`
markers plus per-cluster silhouette quality; its plot shows the selection
curve, the per-group silhouette widths, and the group sizes for both
clustering steps. The state and trajectory rows use distinct palettes, so
the two groupings are not read as corresponding.

```{r evaluate-auto, fig.height = 7}
evaluation <- evaluate(fit_auto)
evaluation
plot(evaluation)
```

## The chapter analysis, explicit where it matters

The chapter's substantive choices stay in one readable call. The state
method defaults to LPA with the chapter's `"EEI"` model, and the three
labels fix the three profiles used from here on; automation only fills
what is omitted.

```{r fit}
fit <- vasstra(
  engagement,
  state_labels = c("Disengaged", "Average", "Active"),
  dissimilarity = "lcs",
  cluster_method = "ward.D2",
  trajectory_labels = c(
    "Mostly active",
    "Mostly average",
    "Mostly disengaged"
  ),
  positive_states = "Active",
  negative_states = "Disengaged",
  seed = 22294
)
fit
```

```{r evaluate-explicit, fig.height = 7}
plot(evaluate(fit))
```

## Tidy fit indices

The fitted LPA solution as one tidy row with the complete
information-criterion family, then the comparison against all candidate
counts, then the trajectory clustering.

```{r fit-indices}
fit_indices(fit)
fit_indices(fit, compare = TRUE)
fit_indices(fit, step = "trajectories", compare = TRUE)
```

## Relabel after inspection

`set_labels()` renames groups in place — full vectors or partial named
renames — and every downstream table follows.

```{r set-labels}
fit_renamed <- set_labels(
  fit,
  states = c("Average" = "Moderate"),
  trajectories = c("Thriving", "Coasting", "Struggling")
)
as.data.frame(fit_renamed, unit = "trajectory")
```

## Tidy outputs

```{r tidy-outputs}
head(as.data.frame(fit, unit = "subject"))
head(as.data.frame(fit, unit = "observation"))
head(as.data.frame(fit, unit = "state_profile"))
as.data.frame(fit, unit = "trajectory")
```

## State overview: profile + bars + heatmap + sizes

`type = "all"` draws the four state views together on the shared palette;
the profile heatmap reads with the first state as its top row, matching the
bars, sizes, and legends.

```{r state-all, fig.height = 8}
plot(fit, which = "states", type = "all")
```

Each view is also available alone:

```{r state-profile}
plot(fit, which = "states", type = "profile")
```

```{r state-bars}
plot(fit, which = "states", type = "bars")
```

```{r state-heatmap}
plot(fit, which = "states", type = "heatmap")
```

```{r state-sizes}
plot(fit, which = "states", type = "sizes")
```

## Palette and state order

`colors` recolours a single plot; the states are coloured by position, so
supply one colour per state in state order.

```{r palette-oneoff}
plot(fit, which = "states", type = "sizes",
     colors = c("#D55E00", "#0072B2", "#009E73"))
```

To reuse a palette everywhere without repeating `colors`, pin it on the fit
with `state_colors` (a named vector is matched by state and survives
reordering and `set_labels()`). `state_order` fixes the order the states
read in every plot — stacking, legends, transition and flow nodes —
independent of the order the clusters were discovered in.

```{r palette-stored, fig.height = 5.5}
fit_styled <- vasstra(
  engagement,
  state_labels = c("Disengaged", "Average", "Active"),
  state_order  = c("Active", "Average", "Disengaged"),
  state_colors = c(Disengaged = "#D55E00", Average = "#0072B2", Active = "#009E73"),
  dissimilarity = "lcs",
  cluster_method = "ward.D2"
)
plot(fit_styled, type = "distribution")
```

## Choice comparisons across methods

The explicit comparison draws a metric-aware plot, on the shared palette
with light gridlines.

```{r state-choices}
state_options <- state_choices(
  engagement,
  n_states = 2:5,
  method = c("kmeans", "pam", "ward.D2", "lpa"),
  lpa_model = "EEI",
  seed = 22294
)
state_options
plot(state_options, metric = "silhouette")
plot(state_options, metric = "bic", groups = "lpa/EEI")
```

## All sequences: heatmap (default), distribution, index

Every plot containing state sequences is rendered by
`Nestimate::sequence_plot()`. The heatmap is the default view for the
complete sequence set.

```{r sequences-heatmap}
plot(fit, which = "sequences")
```

```{r sequences-distribution}
plot(fit, which = "sequences", type = "distribution")
```

```{r sequences-index}
plot(fit, which = "sequences", type = "index")
```

## Flow plots (cograph)

`flow_plot()` renders state movement between consecutive time points
through cograph, on the same palette and state order as every other plot,
and returns a `ggplot` object rather than drawing base graphics. Aggregated
bands and individual lines are the two views:

```{r flow-alluvial, eval = has_cograph}
flow_plot(fit)
```

```{r flow-individual, eval = has_cograph}
flow_plot(fit, type = "individual")
```

The `vignette("flow-plots")` article covers the rest — colouring by
destination or by first/last state, restricting to one trajectory,
line bundling, thresholding negligible flows, and tuning a figure for a
manuscript.

## Transition network (Nestimate builds, cograph draws)

Node size is a transition-network centrality from
`Nestimate::net_centrality()`; the network comes from
`Nestimate::build_tna()`. Drawn by `cograph::splot()`; returns the
tidy node table invisibly. The paired form places the sequences beside the
network on one shared palette.

```{r transition-default, fig.height = 6, eval = has_cograph}
transition_plot(fit)
```

```{r transition-paired, fig.width = 11, fig.height = 4.6, eval = has_cograph}
transition_plot(fit, sequences = TRUE)
```

The centralities are also available as a tidy table:

```{r transition-centrality}
transition_centrality(fit)
```

See `vignette("flow-plots")` for the centrality-driven node sizing
(`size`, `weights`, `loops`), the per-trajectory networks, and the other
paired views.

## Trajectories: distribution, index, heatmap

```{r trajectories-distribution}
plot(fit, type = "distribution")
```

```{r trajectories-index}
plot(fit, type = "index")
```

```{r trajectories-heatmap}
plot(fit, type = "heatmap")
```

## The combined per-trajectory grid

Passing several views in `type` composes them into one figure with **one
row per trajectory and one column per view**, in the order requested — the
standard per-cluster VaSSTra display, drawn in a single call with a shared
ordered legend. The transition column needs the suggested `cograph`
package.

```{r trajectories-grid, fig.width = 11, fig.height = 8, eval = has_cograph}
plot(fit, type = c("transition", "index", "distribution"))
```
