---
title: "VaSStra Tutorial: Variables, States, Sequences, Trajectories"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{VaSStra Tutorial: Variables, States, Sequences, Trajectories}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  dpi = 120
)
```

Longitudinal engagement data have a structural problem: eight indicators
per student per course are too many dimensions to compare students on,
and averaging them across time destroys exactly the thing that matters —
how each student changes. The VaSSTra method solves this in three
compressions. Multivariate observations become a small set of
interpretable **states**; each student's states, ordered in time, become
a **sequence**; and students whose sequences resemble each other become
**trajectories** — a person-centered summary of change that a single
regression coefficient cannot express.

This tutorial runs the complete workflow on the longitudinal engagement
data from the
[VaSSTra chapter](https://lamethods.org/book1/chapters/ch11-vasstra/ch11-vasstra.html)
of *Learning Analytics Methods and Tutorials*, in this order:

1. The complete analysis in one call.
2. Choosing and defending the number of states.
3. Understanding what the states are.
4. Sequences: how states unfold over time.
5. Trajectories: groups of similar journeys.
6. Describing what the trajectories mean.
7. Naming groups after inspecting them.
8. Full control: explicit steps and grid comparisons.

| Task | Verb |
|---|---|
| Complete analysis | `vasstra()` |
| Compare cluster counts | `evaluate()`, `plot()` |
| Fit statistics | `fit_indices()` |
| Rename groups | `set_labels()` |
| Tidy results | `as.data.frame()`, `summary()` |
| Individual steps | `step1_states()` … `step4_describe()` |
| Explicit grids | `state_choices()`, `trajectory_choices()` |

## 1. The complete analysis in one call

The `engagement` data ship analysis-ready: 142 students at eight course
positions, with eight course-standardized indicators and attached role
metadata, so no column plumbing is needed. The chapter's substantive
decision — three latent profiles named Disengaged, Average, and Active —
is expressed entirely by the labels: three labels mean three states,
estimated by latent profile analysis (mclust `"EEI"`, the chapter's
tidyLPA Model 1) because that is the default state method. The LCS
distance with Ward clustering follows the chapter's trajectory analysis.

```{r one-call}
library(VaSStra)
data("engagement", package = "VaSStra")

fit <- vasstra(
  engagement,
  state_labels = c("Disengaged", "Average", "Active"),
  dissimilarity = "lcs",
  cluster_method = "ward.D2",
  positive_states = "Active",
  negative_states = "Disengaged"
)
fit
```

The one remaining automated decision — the number of trajectories — is
reported in the message and stored with its full comparison table in the
fit, so it can be defended later rather than taken on faith. Leaving the
labels out too (`vasstra(engagement)`) automates the state count as
well; automation never selects a solution whose smallest group holds
under five percent of observations, the conventional threshold below
which latent classes are treated as spurious.

## 2. Choosing and defending the number of states

A fitted clustering is a claim, and the claim needs two kinds of
evidence: that the chosen count beats its neighbors, and that the chosen
solution is internally sound. `evaluate()` provides both in one object —
a candidate table across counts and a per-group quality table for the
fitted solution.

```{r evaluate}
evaluation <- evaluate(fit)
evaluation
```

```{r evaluate-plot, fig.height = 6.5}
plot(evaluation)
```

The two rows of panels answer different questions. The trajectory
selection curve peaks at the fitted three groups (silhouette 0.51
against 0.44 and 0.40 for two and four), so the automated choice is also
the local optimum — the strongest position a selection can be in. The
state curve instead *declines* from two: with LPA the count was fixed by
the three labels, and the panel makes the cost of that substantive
choice visible (0.26 at three against 0.35 at two) instead of hiding it.
The silhouette-by-group bars then locate where cohesion comes from: the
middle trajectory is the most compact (0.56), while the smallest one
(0.36) is the first place to look when judging whether three groups
over-merge distinct journeys.

For the model-based evidence reviewers expect, `fit_indices()` returns
the complete information-criterion family as one tidy row — the same
content as tidyLPA's fit table, plus the classification-quality columns
that usually have to be assembled by hand.

```{r fit-indices}
fit_indices(fit)
```

Three of these numbers carry the interpretation. Entropy (0.89) says the
posterior classification is sharp overall; `prob_min` (0.94) says even
the *worst* class assigns its members with high confidence — entropy
summarizes the whole matrix, `prob_min` guards its weakest class, and
reporting both prevents a good average from hiding a bad class. The
criterion family (AIC through ICL) is there for the comparison table:

```{r fit-indices-compare}
fit_indices(fit, compare = TRUE)
```

Every criterion keeps falling through six classes — a nearly linear
decline with no elbow — which is precisely why the chapter's three
profiles rest on interpretability and the classification-quality
columns rather than on a BIC minimum. The table makes that argument
honest: the `best` flag marks what pure criteria would pick, `fitted`
marks what was chosen, and the two need not agree.

## 3. Understanding what the states are

The states are only useful if they can be named, and naming requires
knowing how the profiles differ. The overview draws the four state views
together; each answers one question.

```{r state-all, fig.height = 7.5}
plot(fit, which = "states", type = "all")
```

The profile lines answer *what makes each state itself*: Active sits
around one standard deviation above course average on every indicator,
Disengaged about one below, Average on the line — engagement here is a
level, not a trade-off between, say, forum and lecture activity, which
is exactly what justifies ordered labels. The bars carry the same means
but grouped by indicator, which is the view for spotting whether any
single indicator breaks the ordering (none does). The heatmap adds the
exact magnitudes the lines only sketch — Disengaged is most extreme on
session count (−1.23) and least on lecture views (−0.85). The sizes
panel grounds prevalence: Average is the plurality (45%), and no state
is rare enough to question.

## 4. Sequences: how states unfold over time

The default sequence view is the heatmap: every aligned sequence as one
row, rows ordered by similarity, so persistence shows as long unbroken
bands and instability as speckle — the one view that keeps all 142
individual journeys visible at once.

```{r sequences-heatmap, fig.height = 5.5}
plot(fit, which = "sequences")
```

Two further views split the temporal question the heatmap compresses.
The distribution answers the *aggregate* question — does the
composition of the cohort drift over the eight positions — while the
index plot answers the *individual* one on a fixed subject order. The
distinction matters because a flat distribution can hide heavy
individual churn.

```{r sequences, fig.height = 5.5}
plot(fit, which = "sequences", type = "distribution")
plot(fit, which = "sequences", type = "index")
```

## 5. Trajectories: groups of similar journeys

Grouping the sequences (LCS distance, Ward linkage) yields three
trajectories of 41, 73, and 28 students. The grouped index plot is the
core evidence that the groups mean something: one group is dominated by
Active rows throughout, one by Disengaged, and the largest mixes Average
with excursions — visibly different journeys, not arbitrary partitions
of similar ones.

```{r trajectories, fig.height = 5.5}
plot(fit, type = "index")
plot(fit, type = "distribution")
```

## 6. Describing what the trajectories mean

The per-trajectory summary turns those pictures into numbers, one row
per trajectory.

```{r describe}
as.data.frame(fit, unit = "trajectory")
```

The two time-weighted columns exist because `positive_states` and
`negative_states` were declared in the fit: integrative potential is the
late-weighted share of time in Active, negative exposure the same for
Disengaged. Here they separate the groups almost completely (0.77 versus
0.11 versus 0.00 for integrative potential) — later time points count
more, so these indices read as *where the student is heading*, which
raw state proportions cannot say. Entropy and transition counts add the
stability dimension: the large middle group is not only mixed in state
but also the most mobile.

Per-student indices for downstream modeling — one row per student with
trajectory membership, entropy, complexity, and both exposure indices —
come from the same fit:

```{r indices}
head(as.data.frame(fit))
```

## 7. Naming groups after inspecting them

Interpretation happens *after* fitting, so labels should be assignable
afterward without refitting. `set_labels()` renames in place — every
table, sequence, and recorded positive/negative state follows, and no
fitted value changes. Full vectors rename everything; named renames
change only what they name.

```{r set-labels}
fit <- set_labels(
  fit,
  trajectories = c("Mostly active", "Mostly average", "Mostly disengaged")
)
summary(fit)
```

## 8. Full control: explicit steps and grid comparisons

Everything above is also four explicit steps, each with the same
automated defaults — the route to take when intermediate objects are
needed, when states come from another method (enter at
`step2_sequences()` with a state column), or when the method itself is
under comparison:

```{r steps, eval = FALSE}
states <- step1_states(engagement, n_states = 3)
sequences <- step2_sequences(states)
trajectories <- step3_trajectories(sequences, dissimilarity = "lcs",
                                   method = "ward.D2")
description <- step4_describe(trajectories, positive_states = "Active",
                              negative_states = "Disengaged")
```

When the clustering method is itself in question, `state_choices()` and
`trajectory_choices()` fit the full grid and return one tidy row per
candidate — nothing is selected silently, recommendations are marked
within each method (silhouette for hard clustering, BIC for LPA, since
model-based criteria only compare within LPA), and the inspected
candidate is fitted by describing it:

```{r choices, fig.height = 5}
state_options <- state_choices(engagement, n_states = 2:4,
                               method = c("lpa", "kmeans"))
state_options
plot(state_options, metric = "silhouette")
states <- fit_state_choice(state_options, n_states = 3, method = "lpa",
                           labels = c("Disengaged", "Average", "Active"))
```

The choice plot and `evaluate()` show the same silhouette on the same
scale, so the method comparison here and the count comparison there
compose into one argument. Method overview and formal background:
the [VaSSTra chapter](https://lamethods.org/book1/chapters/ch11-vasstra/ch11-vasstra.html).
